Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
mostly minor changes, but also shortened a lot of overly long lines in the latex source

One key to SciGaP is its openness. While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
stakeholder requirements. The project is leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways into commodity services.

The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as
a ``body of persons of common and especially professional interests scattered through a larger society.'' This makes support
challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging particular types of use.

@@

Software sustainability relies on contributions from all sectors of user community. Together SciPy and NumFOCUS support these sectors. By maximizing contributions growing the user base they help develop and mature Python.

The L\"{o}ffler paper describes the Cactus project which was started in 1996 by participants in USA Binary Black Hole Alliance
Challenge. Cactus has a flesh and thorns model - a community-oriented framework that allows researchers to easily work
together with reusable and extendable software elements. Modules are compiled into an executable and can remain dormant, becoming
active only when parameters and simulation data dictate. Users can  use modules written by others or can write their own
modules without changing other code. The community has grown and diversified beyond the original science areas.

@@

The Wilkins-Diehr paper describes an NSF software institute effort to build a community of those creating science gateways or
web portals for science. These gateways, as described in some of the other papers in this section, can be quite capable, having a
remarkable impact on science.

This paper mentioned challenges highlighted by other papers in this section - mainly the conflict between funding for research
vs infrastructure and the challenges around getting academic credit for infrastructure. Because the authors have studied
many projects, they've also observed how development is often done in an isolated hobbyist environment. Developers are unable
to take advantage of similar work done by others. isolation even when projects have common needs.
But often projects struggle for sustainable funding because they provide infrastructure to conduct
research and many times only the research is funded. Gateways also may start as small group research project, taking off in popularity
once they go live without any long term plans for sustainability or without resources in the project to plan for such.
Subsequent disruptions in service can limit effectiveness and test the limits of the research community's trust. The impact
of gateways can be increased substantially if we understand what makes them successful.

Recommendations from an early study of successful gateways include the following. Leadership and management teams should
design governance to represent multiple strengths and perspectives, plan for change and turnover in the future,
recruit a development team that understands both the technical and domain-related issues, consider sustainability and measure success early and often.
Successful projects recognize the benefits and costs of hiring a team of professionals, demonstrate credibility through stability and
clarity of purpose, leverage the work of others and plan for flexibility.
Successful projects identify an existing community and understand that communities' needs before beginning. Projects adapt
as the communities' needs evolve. For funders, the lifecycle of technology projects must be considered. Solicitations should
be designed to reward effective planning, recognize the benefits and limitations of both technology innovation and reuse,
expect adjustments during the production process, copy effective models from other industries and sectors, and encourage partnerships that support gateway sustainability.

>>>>

One key to SciGaP is its openness. While based on the Apache Airavata project
and the CIPRES Workbench Framework, community contributions are encouraged
because of its open source, open governance and open operations policies. The
goal is robust, sustainable infrastructure with a cycle of development that
improves reliability and prioritizes stakeholder requirements. The project is
leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways
into commodity services.

The Zentner paper describes experiences and challenges with the large
nanoHUB.org community. The authors define community as a ``body of persons of
common and especially professional interests scattered through a larger
society.'' This makes support challenging because of the diversity of
viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging
particular types of use.

@@

Software sustainability relies on contributions from all sectors of the user
community. Together SciPy and NumFOCUS support these sectors. By maximizing
contributions growing the user base they help develop and mature Python.

The L\"{o}ffler paper describes the Cactus project which was started in 1996 by
participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
and thorns, core and module, model - a community-oriented framework that allows
researchers to easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain dormant,
becoming active only when parameters and simulation data dictate. Users can
use modules written by others or can write their own modules without changing
other code. The community has grown and diversified beyond the original science
areas.

@@

The Wilkins-Diehr paper describes an NSF software institute effort to build a
community of those creating science gateways or web portals for science. These
gateways, as described in some of the other papers in this section, can be
quite capable, having a remarkable impact on some parts of science.

This paper mentioned challenges highlighted by other papers in this section -
mainly the conflict between funding for research vs infrastructure and the
challenges around getting academic credit for infrastructure. Because the
authors have studied many projects, they've also observed how development is
often done in an isolated hobbyist environment. Developers are unable to take
advantage of similar work done by others, isolation even when projects have
common needs. But often projects struggle for sustainable funding because they
provide infrastructure to conduct research and many times only the research is
funded. Gateways also may start as small group research project, taking off in
popularity once they go live without any long term plans for sustainability or
without resources in the project to plan for such. Subsequent disruptions in
service can limit effectiveness and test the limits of the research community's
trust. The impact of gateways can be increased substantially if we understand
what makes them successful.

Recommendations from an early study of successful gateways include the
following. Leadership and management teams should design governance to
represent multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the technical and
domain-related issues, consider sustainability and measure success early and
often. Successful projects recognize the benefits and costs of hiring a team of
professionals, demonstrate credibility through stability and clarity of
purpose, leverage the work of others and plan for flexibility. Successful
projects identify an existing community and understand the communities' needs
before beginning. Projects adapt as the communities' needs evolve. For funders,
the lifecycle of technology projects must be considered. Solicitations should
be designed to reward effective planning, recognize the benefits and
limitations of both technology innovation and reuse, expect adjustments during
the production process, copy effective models from other industries and
sectors, and encourage partnerships that support gateway sustainability.
